Determination of Center of Rotation \\ in Computed Tomography

### 2017-07-19 seminar_cor/seminar_cor_haegele.tex
wrote center of mass section

\section{COR Determination}
The problem of incorrect assumption of the location of the COR can occur in CT setups with misaligned source and detector.
Especially in industrial CT with resolutions on the micrometer scale, calibrating a CT system correctly can be very challenging.
In parallel beam geometry, a misaligned COR, i.e. the COR is projected onto $\tau$ on the $r$-axis, results in a shift of the sinogram along the $r$-axis by $\tau$.
\begin{equation}
R_\tau(r,\gamma) = R(r-\tau,\gamma)
\end{equation}
The actual problem arises when reconstructing the signal from $R_\tau$ as the algorithms, e.g. filtered back projection (FBP), assume the COR to be in the center.
Thus, ray intersections do not lign up correctly and the reconstructed image suffers from streaking artifacts.
\Cref{fig:shiftfbp} shows the effect of shift of COR in FBP.
%
\begin{figure}[!h]
\centering
\subfloat[$\tau=0$px]{\includegraphics[width=0.24\linewidth]{img/fbp_shift_0}}
\hfil
\subfloat[$\tau=2$px]{\includegraphics[width=0.24\linewidth]{img/fbp_shift_2}}
\hfil
\subfloat[$\tau=4$px]{\includegraphics[width=0.24\linewidth]{img/fbp_shift_4}}
\hfil
\subfloat[$\tau=8$px]{\includegraphics[width=0.24\linewidth]{img/fbp_shift_8}}
\caption{
FBP of the phantom from \cref{fig:sinogram_parallel} with different shifts ($\tau$) of COR.
The resolution of the sinogram in $r$ direction was 256 pixels.
With incresing shift, stronger vertical streaking artifacts are introduced into the reconstructed image.
}
\label{fig:shiftfbp}
\end{figure}
%

\subsection{Sinogram Based Methods}

\subsubsection{High Density Feature}

\subsubsection{Center of Mass}

>>>>

\section{COR Determination}
The problem of incorrect assumption of the location of the COR can occur in CT setups with misaligned source and detector.
Especially in industrial CT with resolutions on the micrometer scale, calibrating a CT system correctly can be very challenging.
In parallel beam geometry, a misaligned COR, i.e. the COR is projected onto $\tau$ on the $r$-axis, results in a shift of the sinogram along the $r$-axis by $\tau$.
\begin{equation}
\label{eq:radonoffset}
R_\tau(r,\gamma) = R(r-\tau,\gamma)
\end{equation}
The actual problem arises when reconstructing the signal from $R_\tau$ as the algorithms, e.g. filtered back projection (FBP), assume the COR to be in the center.
Thus, ray intersections do not lign up correctly and the reconstructed image suffers from streaking artifacts.
\Cref{fig:shiftfbp} shows the effect of shift of COR in FBP.
%
\begin{figure}[!h]
\centering
\subfloat[$\tau=0$px]{\includegraphics[width=0.24\linewidth]{img/fbp_shift_0}}
\hfil
\subfloat[$\tau=2$px]{\includegraphics[width=0.24\linewidth]{img/fbp_shift_2}}
\hfil
\subfloat[$\tau=4$px]{\includegraphics[width=0.24\linewidth]{img/fbp_shift_4}}
\hfil
\subfloat[$\tau=8$px]{\includegraphics[width=0.24\linewidth]{img/fbp_shift_8}}
\caption{
FBP of the phantom from \cref{fig:sinogram_parallel} with different shifts ($\tau$) of COR.
The resolution of the sinogram in $r$ direction was 256 pixels.
With incresing shift, stronger vertical streaking artifacts are introduced into the reconstructed image.
}
\label{fig:shiftfbp}
\end{figure}
%

In order to correctly reconstruct the signal, the COR first need to be determined.
Several approaches to this problem can be found in the literature, which will be introduced in the following subsections.
The variety of approaches can subdivided in sinogram base methods, which analyze a sinogram in order to find the COR, and reconstruction based methods, which compare reconstruction results with different COR in an iterative fashion.

\subsection{Sinogram Based Methods}
In this subsection the approaches that are based on sinogram analysis are introduced.
The approaches are presentred in chronical order.

\subsubsection{High Density Feature}
Crawford~et~al.~\cite{crawford1988} presented an approach for finding the COR in which they used two $180^\circ$ opposing projections of a distinct high density feature.
When locating this feature in the two projections, the COR can be calculated as the sum of the detector coordinates of the feature divided by two.
This becomes clear when thinking about the feature being offset from the COR by $\delta$ in projection $p_1$.
The corresponding position of the feature in projection $p_2$ will then be on the opposite side of the COR, offset by $-\delta$ from the COR. \Cref{fig:highdenistyfeature} illustrates this concept.
%
\begin{figure}[!h]
\centering
\subfloat[phantom]{\includegraphics[width=0.32\linewidth]{img/highdensity_inv.png}}
\hfil
\subfloat[sinogram]{\includegraphics[width=0.65\linewidth]{img/highdensityradon_degrees_feature_centerline_horiz.png}}
\caption{
A phantom with a high density feature is shown on the left side.
In corresponding sinogram on the right, the sinusodial trace of the feature can be observed.
The dotted line indicates the COR in the sinogram.
It can be seen that the feature locations in $180^\circ$ opposing projections have the same distance to the COR.  
}
\label{fig:highdenistyfeature}
\end{figure}
%

This approach could also be expanded to use all projections in the sinogram.
For each opposing pair of projections the COR could be calcualted and then the mean of all CORs will yield the final result.
This approach can be automated if there esists a distinct global maximum in the projections of the sinogram, otherwise a location of the correct feature is not guaranteed.

\subsubsection{Center of Mass}
Azevedo~et~al.~\cite{azevedo90} proposed a more generally applicable algorithm, in which they replaced the high density feature from \cite{crawford1988} with the center of mass.
The center of mass of the original signal is a point $(\overline{x},\overline{y}$) which's coordinates can be calculated as follows (calculation of $\overline{y}$ analogous to $\overline{x}$).
%
\begin{equation}
\overline{x} = \frac{
\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}x\;f(x,y)\;dx\;dy
}
{
\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}f(x,y)\;dx\;dy
}
\end{equation}
%
As the center of mass of the original signal $f(x,y)$ is projected onto the projection center of mass in parallel beam geometry, it can be used as a reference feature.
The projection center of mass is dependent on the projection angle and can be calculated as follows.
%
\begin{equation}
\overline{r_\tau}(\gamma) = \frac{
\int_{-\infty}^{\infty}r\;R_\tau(r,\gamma)\;dr
}
{
\int_{-\infty}^{\infty}R_\tau(r,\gamma)\;dr
}
=
\frac{
\int_{-\infty}^{\infty}r\;R(r-\tau,\gamma)\;dr
}
{
\int_{-\infty}^{\infty}R_(r-\tau,\gamma)\;dr
}
\end{equation}
%
where $R_\tau(r,\gamma)$ is the radon transform with COR offset by~$\tau$ (\cref{eq:radonoffset}).
The previous equation can be simplified to the following.
%
\begin{equation}
\label{eq:projection_centerofmass}
\overline{r_\tau}(\gamma) = \tau + \overline{r}(\gamma) = \tau + \overline{x}\;cos(\gamma) + \overline{y}\;sin(\gamma)
\end{equation}
%
Using linear regression (least squares) the parameters $\tau$, $\overline{x}$ and $\overline{y}$ can be approximated from the matrix notation of \cref{eq:projection_centerofmass}.
%
\begin{equation}
\begin{pmatrix}
1 & cos(\gamma_1) & sin(\gamma_1) \\
\vdots & \vdots & \vdots \\
1 & cos(\gamma_n) & sin(\gamma_n) 
\end{pmatrix}
\cdot
\begin{pmatrix}
\tau \\
\overline{x} \\
\overline{y}
\end{pmatrix}
=
\begin{pmatrix}
\overline{r_\tau}(\gamma_1) \\
\vdots \\
\overline{r_\tau}(\gamma_n)
\end{pmatrix}
\end{equation}
%
Even though $\overline{x}$ and $\overline{y}$ are not needed for a correct reconstruction, it is a neat bonus.
